\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to express my deepest gratitude to all those who have supported and guided me throughout this doctoral journey.

First and foremost, I am profoundly grateful to my supervisors, Professor Alexander Gennadevich Trofimov and Professor Yuri Vladimirovich Borodachev, for their invaluable guidance, encouragement, and unwavering support throughout this research. Their expertise, insights, and constructive feedback have been instrumental in shaping this work. I am particularly thankful for their patience and understanding during challenging moments of the research process.

I would like to thank the faculty and staff at the National Research Nuclear University MEPhI, particularly those at the Artificial Intelligence Research Center, for providing an excellent academic environment and research facilities. The intellectual atmosphere and collaborative spirit at MEPhI have been crucial to the successful completion of this dissertation.

My sincere appreciation extends to my colleagues and fellow researchers who have provided valuable discussions, feedback, and collaboration opportunities. Special thanks to the participants of various conferences and workshops where portions of this work were presented, whose questions and suggestions helped refine and improve the research.

I am grateful to the reviewers and examiners of this dissertation for their time, effort, and constructive criticism, which has undoubtedly strengthened the quality of this work.

I acknowledge the support provided through various research grants and computing resources that made the extensive experimental validation possible. Access to high-performance computing facilities and datasets was essential for conducting the comprehensive evaluations presented in this thesis.

I would like to express my heartfelt thanks to my family and friends for their continuous encouragement, patience, and understanding throughout the years of doctoral study. Their emotional support and belief in my abilities have been a constant source of motivation.

Finally, I am grateful to the broader research community working on machine learning, cybersecurity, and network intrusion detection. Standing on the shoulders of giants, this work builds upon the foundational contributions of numerous researchers whose dedication to advancing knowledge has paved the way for this investigation.

\vspace{2cm}

\begin{flushright}
Roger Nick Anaedevha\\
Moscow, November 2025
\end{flushright}

\cleardoublepage
